---
title: "r4ds-ch16"
---

FACTORS --> `forcate` package in `tidyverse`

```{r}
library(tidyverse)
```

`forcate` package, part of `tidyverse` core, provides tools for dealing with categorical variables

```{r}
x1 <- c("Dec", "Apr", "Jan", "Mar")  # don't sort alphabetically, not what we want
sort(x1)
```

```{r}
x2 <- c("Dec", "Apr", "Jam", "Mar")  # Ops. Typo & don't sort alphabitically, not what we want
sort(x2)
```

## Create factor

Step 1: create a list
```{r}
month_levels <- c(
  "Jan", "Feb", "Mar", "Apr", "May", "Jun",
  "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"
)
```

Step 2: Use the list to create a factor
```{r}
y1 = factor(x1, levels = month_levels)
sort(y1)
```

Values not in the list will be silently converted to `NA` which is risky
```{r}
y2 <- factor(x2, levels = month_levels)
y2
```

So, instead we use `forcat::fct`
```{r}
y2.2 <- forcats::fct(x2, levels = month_levels)
```


When removing the `levels` argument, `factor` sort levels alphabetically but `forcats::fct` sort them by first appearance
```{r}
y1 <- factor(x1)
sort(y1)
levels(y1)
```

```{r}
y1 <- forcats::fct(x1)
y1
levels(y1)
```

## `forcats::gss_cat` dataset

A sample of data from the General Social Survey, a long-running US survey collected by the independent research organization NORC at the University of Chicago.  The survey has thousands of questions but `gss_cat` has a handful that will illustrate some common challenges you’ll encounter when working with factors.

```{r}
gss_cat
```

```{r}
?gss_cat
```

Look at levels of some factors
```{r}
gss_cat |> 
  count(race)
```

```{r}
summary(gss_cat)
```

## Modify Factor Order

```{r}
gss_cat |> 
  group_by(relig) |> 
  summarise(
    tvhours = mean(tvhours, na.rm = TRUE),
    n = n()
  ) |> 
  ggplot(aes(x = tvhours, y = fct_reorder(relig, tvhours, .desc = TRUE))) + 
  geom_point()
```

## Move Level to Front
```{r}
gss_cat |> 
  group_by(rincome) |> 
  summarise(
    age = mean(age, na.rm = TRUE),
    n = n()
  ) |> 
  ggplot(aes(x = age, y = fct_relevel(rincome, "Not applicable"))) + 
  geom_point()
```

## Reorder Levels based on Ending Value of Anohter

This allow the lines of the different levels to align with legend

```{r}
gss_cat |> 
  filter(!is.na(age)) |> 
  count(age, marital) |> 
  group_by(age) |> 
  mutate(
    prop = n / sum(n)
  ) |> 
  ggplot(aes(x = age, y = prop, color = fct_reorder2(marital, age, prop))) + 
  geom_line()
```

## Reorder Levels for Bar Plots
```{r}
gss_cat |> 
  mutate(marital = marital |> fct_infreq() |> fct_rev()) |> 
  ggplot(aes(marital)) + 
  geom_bar()
```


## Modifying Factor Levels

- `fct_recode`
- `fct_collapse` --> multiple groups to one
- `fct_lump*()` --> lump together small groups
   - `fct_lump_lowfreq()` --> lump the smallest groups into one
   - `fct_lump_n()` --> lump into n groups
   - `fct_lump_min()`
   - `fct_lump_prop()`

## Order Factors

There is order but the distance between levels in unknown.  Notice the prnt.

```{r}
ordered(c("a", "b", "c", "d"))
```

## Resouces

- [`forcats` Reference](https://forcats.tidyverse.org/reference/index.html)
- [Wrangling categorical data in R](https://peerj.com/preprints/3163/) by Amelia McNamara and Nicholas Horton to learn more about factors.
